---
title: "Lab 3"
author: "Aron, Sergey, Shahin"
date: "2025-12-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE 
)

```
## Statement of Contribution

Assignment 2 was mainly contributed by Aron. 
Assignment 3 was mainly contributed by Shahin.
Assignment 4 was mainly contributed by Sergey. 
Theory questions: Q1: Aron, Q2: Shahin, Q3: Sergey.

## Assignment 1. THEORY

**Q1: What is the kernel trick?** Page 194: If x is used in the model only as dotproducts, we can use the kernel for phi(x)phi(x)^T. The kernel could be easier to calculate, the polynomial basis has closed form so we do not need to project into higher-dimensional space.

**Q2: In the literature, it is common to see a formulation of SVMs that makes use of a hyperparameter
C. What is the purpose of this hyperparameter?**

**Q3: In neural networks, what do we mean by mini-batch and epoch?** Page 125: A *mini-batch* is a small subset of the full dataset, typically containing $n_b = 10$, $n_b = 100$, or $n_b = 1{,}000$ observations.

A single complete pass through the entire training dataset is referred to as an *epoch*. Each epoch therefore consists of $n / n_b$ iterations.


## Assignment 2. KERNEL METHODS

```{r 2_1}
#######################
#
# ASSIGNMENT 2 CODE
#
#######################
set.seed(1234567890)
library(geosphere)
stations <- read.csv("data/stations.csv", fileEncoding = "latin1")
temps <- read.csv("data/temps50k.csv")
st <- merge(stations,temps,by="station_number")
h_distance <- # These three values are up to the students
h_date <-
h_time <-
#this is linköping
latitude <- 58.4274 # The point to predict (up to the students)
longitude <- 14.826

# documentation says first one is longitude
# distance is in meters, i checked externally
distance_to_station <- sapply(1:nrow(st), function(x) distHaversine(c(st$longitude[x], st$latitude[x]), c(longitude, latitude)))

st$distance <- distance_to_station
```



```{r 2_2}


st$date <- st$date |> as.Date()


# christmas 2008
date_of_interest <- "2008-12-24" |> as.Date()

times <- seq(4, 24, by = 2) |> as.character()
times <- sapply(1:length(times), 
                function(x) {
  if(nchar(times[x]) == 1){return(paste0("0", times[x]))} 
                return(times[x])}
  )
times <- paste0(times, ":00:00")
temp <- vector(length=length(times))
temp_mult <- vector(length=length(times))

for (i in seq_along(times)) {
  hour_of_interest <- times[i]
  numeric_hour_of_interest <- substr(hour_of_interest, 1, 2) |> as.numeric()

date_diff <- difftime(date_of_interest, st$date) |> as.numeric()
st$day_diff <- date_diff 

# negative happen after our date_of_interest
posterior_index1 <- date_diff < 0


# inefficient code but sufficient, if clean use lubridate
same_day <- st$date == date_of_interest 
hour_later <- (substr(st$time, 1, 2) |> as.numeric()) > numeric_hour_of_interest

posterior_index2 <- hour_later & same_day

posterior_index <- posterior_index1 | posterior_index2


prior_df <- st[!posterior_index, ]


# time is cyclic, so if we want to have hour of day as a variable, 
# similar times to 23 are both 21 and 01...
hour_fun <- function(x, y) {
  delta <- abs(x - y)
  
  if (delta <= 12) {
    return(delta)
  } else {
    if (x < y) {
      
      return( abs((x + 24) - y) )
      
    } else {
      return( abs((y + 24) - x) )
    }
    
  }
  
}


prior_df$delta_t <- sapply(1:nrow(prior_df), function(x) hour_fun(numeric_hour_of_interest, prior_df$time[x] |> substr(1, 2) |> as.numeric()))

# small h makes the distance larger
h_distance <- sd(prior_df$distance) * 0.1
h_date <- sd(prior_df$day_diff) * 0.1
h_time <- sd(prior_df$delta_t) * 0.1

# too large h means it converges to the mean
# mean(prior_df$air_temperature)

a <- sum(dnorm(prior_df$distance / h_distance) * prior_df$air_temperature) + sum(dnorm(prior_df$day_diff / h_date) * prior_df$air_temperature) + sum(dnorm(prior_df$delta_t / h_time) * prior_df$air_temperature) 

b <-  sum(dnorm(prior_df$distance / h_distance)) + sum(dnorm(prior_df$day_diff / h_date)) + sum(dnorm(prior_df$delta_t / h_time))

prediction <- a/b


h_distance <- sd(prior_df$distance) * 0.5
h_date <- sd(prior_df$day_diff) * 0.5
h_time <- sd(prior_df$delta_t) * 0.5

c <- dnorm(prior_df$distance / h_distance) * dnorm(prior_df$day_diff / h_date) * dnorm(prior_df$delta_t / h_time) 


prediction_mult <- sum(c * prior_df$air_temperature) / sum(c)

temp[i] <- prediction
temp_mult[i] <- prediction_mult

# loop end
# -------------------
}

plot(temp, type="o")
plot(temp_mult, type="o")

curve(dnorm(x/h_distance), from = 0, to = 1000000)

```


The widths are picked like this, smaller h means we put less impact on points further away. It is scaled by standard deviation, otherwise everything gets pushed to 0 for distance for example. 

h_distance <- sd(prior_df$distance) * 0.1

h_date <- sd(prior_df$day_diff) * 0.1

h_time <- sd(prior_df$delta_t) * 0.1
 
h had to be increased for the multiplicative kernel as it caused division by 0 issues.

The weight from the kernel decays with distance. 

It is worth noting that these predictions are very poor, I tried a data in the summer and got values below 10 degrees. The choice of kernel doesn't handle seasonality but does handle trend effects.

The difference between the multiplicative and the additive kernel is that for the multiplicative kernel the deviation of one variable does not impact the overall kernel that much as for the additive kernel.

## Assignment 3. SUPPORT VECTOR MACHINES

## Assignment 4. NEURAL NETWORKS


```{r t4_setup_and_functions}
#######################
#
# ASSIGNMENT 4 CODE
#
#######################
# Set all variables to null to avoid possible variable name clashes between unrelated parts of the lab
rm(list = ls())

library(neuralnet)
library(sigmoid)

get_data <- function(max_value, train_set_count, total_count=500) {
  set.seed(1234567890)
  Var   <- runif(total_count, 0, max_value)
  mydata <- data.frame(Var, Sin = sin(Var))
  tr <- mydata[1:train_set_count, ] 
  if (total_count > train_set_count) {
    te <- mydata[(train_set_count+1):total_count, ]     
  }

  
  list(tr = tr, te = te)
}

train_network <- function(tr,
                          act_fct   = "logistic",
                          form      = Sin ~ Var,
                          threshold = NULL) {
  set.seed(1234567890)
  winit <- runif(21, min = -1, max = 1)

  args <- list(
    formula       = form,
    data          = tr,
    hidden        = 10,
    startweights  = winit,
    act.fct       = act_fct,
    linear.output = TRUE
  )

  if (!is.null(threshold)) {
    args$threshold <- threshold
  }

  nn <- do.call(neuralnet, args)
  nn
}
```

### Task 1

```{r t4_1}
mydata <- get_data(max_value=10,train_set_count=25)
tr <- mydata$tr
te <- mydata$te
nn <- train_network(tr)

op <- par(mar = c(5, 4, 4, 8), xpd = TRUE)

plot(tr, cex = 3, main = "NN approximation of sin(x) on [0, 10]") 
points(te, col = "blue", cex = 1)                  
points(te[, 1], predict(nn, te), col = "red", cex = 1)

legend("topright",
       inset = c(-0.3, 0),
       legend = c("Train sin(x)", "Test sin(x)", "NN prediction"),
       col    = c("black","blue", "red"),
       pch    = c(1, 1),
       bty    = "n")
```

From the plot above we can see that the neutral network is able to approximate the sine function 
quite well in the input range of 0 to 10 even with the limited amount of training data provided (25 data points).  

### Task 2

```{r t4_2}
h1 <- function(x) x                         # identity
# h2 <- function(x) pmax(0, x)              # ReLU 
# as explained here: 
# https://stackoverflow.com/questions/34532878/
# package-neuralnet-in-r-rectified-linear-unit-relu-activation-function
# it won't work as the pmax function is not differentiable, the suggestion is to use
# the relu function from the sigmoid package
h2 <- relu
h3 <- function(x) log(1 + exp(x))         # softplus


acts <- list(
  linear   = h1,
  ReLU     = h2,
  softplus = h3
)


for (name in names(acts)) {
  h <- acts[[name]]
  
  nn_temp    <- train_network(tr, act_fct = h)
  predictions <- as.vector(predict(nn_temp, te["Var"]))
  y_all       <- c(te$Sin, predictions)
  
  op <- par(mar = c(5, 4, 4, 8), xpd = TRUE)
  
  plot(te$Var, te$Sin,
       col = "blue", cex = 1,
       xlab = "x", ylab = "sin(x)",
       main = paste("NN approximation of sin(x) on [0, 10]\nActivation:", name),
       ylim = range(y_all))
  
  points(te$Var, predictions, col = "red", cex = 1)
  
  legend("topright",
         inset = c(-0.3, 0),
         legend = c("True sin(x)", "NN prediction"),
         col    = c("blue", "red"),
         pch    = c(1, 1),
         bty    = "n")
}
```
From the plots above we can see that the neural network is unable to learn the sine function at all when using the identity function and provides constant output. For the ReLU function, the function is approximated only within a very limited range and as a straight line rather than a curve. The softplus function achieves a potentially better approximation that the default logistic function to which it is closely related. 


### Task 3

```{r t4_3}
set.seed(1234567890)
Var_new  <- runif(500, 0, 50)
newdata  <- data.frame(Var = Var_new, Sin = sin(Var_new))

pred_new <- predict(nn, newdata)   

y_all <- c(newdata$Sin, pred_new)

op <- par(mar = c(5, 4, 4, 8), xpd = TRUE)

plot(newdata$Var, newdata$Sin,
     xlab = "x", ylab = "sin(x)",
     main = "NN approximation of sin(x) on [0, 50]",
     ylim = range(y_all))
points(newdata$Var, pred_new,
       col = "red", pch = 16, cex = 0.7)

legend("topright",
       inset = c(-0.3, 0),
       legend = c("True sin(x)", "NN prediction"),
       col    = c("black", "red"),
       pch    = c(1, 16),
       bty    = "n")
```

From the plot above, we can see that in the range of input values of 0 to 50, the approximation quickly breaks down when the input values exceed 10 and eventually the outputs collapse to a constant.

### Task 4

```{r t4_4}
x_big  <- seq(1, 50, length.out = 200)
dat_big <- data.frame(Var = x_big)
# collapsing to a value around -2.785
W1 <- nn$weights[[1]][[1]]
W2 <- nn$weights[[1]][[2]]
sigm <- function(z) plogis(z)  # logistic activation

# Hidden pre-activations and activations for x
z_hidden <- outer(x_big, W1[2, ], `*`) + matrix(W1[1, ], nrow = length(x_big), ncol = 10, byrow = TRUE)
h_hidden <- sigm(z_hidden)

matplot(x_big, h_hidden, type = "l", lty = 1,
        xlab = "x", ylab = "hidden activations",
        main = "Hidden units for x in [1, 50]")
```
Plotting the outputs of the hidden layer demonstrates that, as the input value increases, the neurons of the hidden layer become "saturated" i.e. more and more of them are stuck at the value of either 0 or 1. When it finally happens for all neurons - on the graph above at a value of approximately 30 - the neural network will produce the same output value regardless of the input value.  

## Task 5

```{r t4_5}
set.seed(1234567890)
Var   <- runif(500, 0, 10)
mydata <- data.frame(Var, Sin = sin(Var))
nn_temp    <- train_network(mydata, form = Var ~ Sin, threshold = 0.1)
predictions <- as.vector(predict(nn_temp, mydata["Sin"]))
y_all       <- c(mydata$Var, predictions)

op <- par(mar = c(5, 4, 4, 8), xpd = TRUE)

plot(mydata$Sin, mydata$Var,
     col = "blue", cex = 1,
     xlab = "sin(x)", ylab = "x",
     main = "NN approximation of sin(x) to x function",
     ylim = range(y_all))

points(mydata$Sin, predictions, col = "red", cex = 1)

legend("topright",
       inset = c(-0.3, 0),
       legend = c("True x", "NN prediction"),
       col    = c("blue", "red"),
       pch    = c(1, 1),
       bty    = "n",
       title  = name)
```
We are seeing a poor approximation because the neural network is trying to find a 1-to-1 mapping between the input and the output values but in the case of the $y=\arcsin(x)$ function it does not exist.


## Appendix


```{r, ref.label=setdiff(knitr::all_labels(), c('setup')), echo=TRUE, eval=FALSE}



```
