---
title: "lab3"
output: html_document
date: "2025-12-09"
---

# Theroy Q1
Page 194: If x is used in the model only as dotproducts, we can use the kernel for phi(x)phi(x)^T. The kernel could be easier to calculate, the polynomial basis has closed form so we do not need to project into higher-dimensional space.  


```{r}
set.seed(1234567890)
library(geosphere)
stations <- read.csv("stations.csv", fileEncoding = "latin1")
temps <- read.csv("temps50k.csv")
st <- merge(stations,temps,by="station_number")
h_distance <- # These three values are up to the students
h_date <-
h_time <-
#this is linköping
latitude <- 58.4274 # The point to predict (up to the students)
longitude <- 14.826

# documentation says first one is longitud
# distance is in meters, i checked externally
distance_to_station <- sapply(1:nrow(st), function(x) distHaversine(c(st$longitude[x], st$latitude[x]), c(longitude, latitude)))

st$distance <- distance_to_station
```



```{r}


st$date <- st$date |> as.Date()


# christmas 2008
date_of_interest <- "2008-12-24" |> as.Date()

times <- seq(4, 24, by = 2) |> as.character()
times <- sapply(1:length(times), 
                function(x) {
  if(nchar(times[x]) == 1){return(paste0("0", times[x]))} 
                return(times[x])}
  )
times <- paste0(times, ":00:00")
temp <- vector(length=length(times))
temp_mult <- vector(length=length(times))

for (i in seq_along(times)) {
  hour_of_interest <- times[i]
  numeric_hour_of_interest <- substr(hour_of_interest, 1, 2) |> as.numeric()

date_diff <- difftime(date_of_interest, st$date) |> as.numeric()
st$day_diff <- date_diff 

# negative happen after our date_of_interest
posterior_index1 <- date_diff < 0


# inefficient code but sufficient, if clean use lubridate
same_day <- st$date == date_of_interest 
hour_later <- (substr(st$time, 1, 2) |> as.numeric()) > numeric_hour_of_interest

posterior_index2 <- hour_later & same_day

posterior_index <- posterior_index1 | posterior_index2


prior_df <- st[!posterior_index, ]


# time is cyclic, so if we want to have hour of day as a variable, 
# similar times to 23 are both 21 and 01...
hour_fun <- function(x, y) {
  delta <- abs(x - y)
  
  if (delta <= 12) {
    return(delta)
  } else {
    if (x < y) {
      
      return( abs((x + 24) - y) )
      
    } else {
      return( abs((y + 24) - x) )
    }
    
  }
  
}


prior_df$delta_t <- sapply(1:nrow(prior_df), function(x) hour_fun(numeric_hour_of_interest, prior_df$time[x] |> substr(1, 2) |> as.numeric()))

# small h makes the distance larger
h_distance <- sd(prior_df$distance) * 0.001
h_date <- sd(prior_df$day_diff) * 0.001
h_time <- sd(prior_df$delta_t) * 0.001

# too large h means it converges to the mean
# mean(prior_df$air_temperature)

K1 <- sum(dnorm(prior_df$distance / h_distance) * prior_df$air_temperature) / sum(dnorm(prior_df$distance / h_distance))

K2 <- sum(dnorm(prior_df$day_diff / h_date) * prior_df$air_temperature) / sum(dnorm(prior_df$day_diff / h_date))

K3 <- sum(dnorm(prior_df$delta_t / h_time) * prior_df$air_temperature) / sum(dnorm(prior_df$delta_t / h_time))

prediction <- mean(c(K1, K2, K3))

x <- (K1*K2*K3)
prediction_mult <- sign(x) * abs(x)^(1/3)

temp[i] <- prediction
temp_mult[i] <- prediction_mult

# loop end
# -------------------
}

plot(temp, type="o")
plot(temp_mult, type="o")

curve(dnorm(x), from = 0, to = 10)

```
The widths are picked like this, smaller h means we put less impact on points further away. It is scaled by standard deviation, otherwise everything gets pushed to 0 for distance for example. 

h_distance <- sd(prior_df$distance) * 0.1
h_date <- sd(prior_df$day_diff) * 0.1
h_time <- sd(prior_df$delta_t) * 0.1


It is worth noting that these predictions are very poor, I tried a data in the summer and got values below 10 degrees. 

The weight from the kernel decays with distance. 
